The Spatial and Temporal Domains of Modern Ecology

### 2017-12-19 paper/SI.tex
More edits to SI on trends section, and additional edits to main text

\subsubsection*{Trends in methods and scale}
To assess the degree to which observational methods changed over the course of the 10 years worth of studies we surveyed, we calculated the percentage of observations made using remote sensing, general field methods, and automated \emph{in situ} methods, and fit a linear regression between these percentages and publication year, weighting the regression by the total number of observations in each year.  We performed the same analysis for the four primary dimensions (resolution, extent, interval, and duration) in order to assess whether there were any trends in observational scales.

\subsubsection*{Extracting and analyzing data from earlier meta-analyses}
To compare the results of our analysis with the observational scales of earlier ecological studies, we used graph capture software\footnote{http://arohatgi.info/WebPlotDigitizer/app/} to extract the data values from Figure 6.1 in Tilman \cite{tilman_ecological_1989}, Figure 1 in Kareiva and Anderson \cite{kareiva_spatial_1988}, and Figure 2 in Porter et al \cite{porter_crop_2005}. To maintain as much comparability as possible with our inclusion criteria, we excluded experiment studies in Tilman's data, as well as the values of any studies having duration values greater than 100 years (no upper time bound was provided for these), leaving duration values for 419 (out of 623) studies. As Tilman presented duration values as a histogram, we calculated the mean duration across all studies as the weighted (by number of observations per bin) mean of bin center-point values (i.e. the weighted mean of the bin means).  We also excluded four (of 29) observation values from Porter et al's data on observational extent and frequency, which, in contrast to the other 25, were not randomly selected. Porter et al also used irregular scales for both their X (frequency) and Y (extent) axis, therefore we had to visually estimate the scale values for each data point after graphical extraction, and converted their extent values (in km) to hectares and their frequency values to intervals. Kareiva and Anderson presented resolution as plot diameters (m), which we squared to make comparable to our resolution metric.

@@

\subsubsection*{Trends in observational methods and scales by year}
\vspace{-10pt}
To evaluate the potential impact that excluding studies from 2015-2017 would have on our findings, we analyzed the trends in 1) ecological observing methods, and 2) typical scales of ecological observations over the 10 year period.

We used linear regression (weighted by the number of observations per publication year) to assess whether the relative frequency of observing methods changed during the 10 year study period. The results (Fig. \ref{type_by_yr}) suggest an increase (by 1.3\% per year) in the use of remote sensing and a corresponding decline in field methods (by the same percentage) during the 2004-2014 time period, although neither trend is statistically significant (remote observation regression, p$<$0.12; field observation regression, p$<$0.18).  Automated sensing methods showed no visible trend.

@@

Of course, we cannot attach confidence to a trend in observational extent extrapolated from a non-significant trend in observational type. However, some additional evidence in support of this possible trend lies within the extent values themselves, which appeared to increase by 0.25 orders of magnitude per year (Fig. \ref{sc_by_yr}) between 2004-2014 (R$^2$ = 0.25, p$<$0.07). This somewhat clearer (albeit still not significant at the customary p$<$0.05 threshold) trend suggests that including more recent studies would have increased mean extent, but by a more modest 5.5\% (0.02 orders of magnitude).

The trend calculations described above can be found in the ``Additional Analyses" vignette in the accompanying \href{https://github.com/agroimpacts/ecoscales}{R package/code repository}.

>>>>

\subsubsection*{Trends in methods and scale}
To evaluate the potential impact that excluding studies from 2015-2017 would have on our findings, we analyzed the trends in 1) ecological observing methods, and 2) typical scales of ecological observations over the 10 year period. To undertake the former assessment, we calculated the percentage of observations made using remote sensing, general field methods, and automated \emph{in situ} methods, and fit a linear regression between these percentages and publication year, weighting the regression by the total number of observations in each year.  For the second analysis, we applied the same regression analysis to the four primary dimensions (resolution, extent, interval, and duration) in order to assess whether there were any trends in observational scales.

The regressions and resulting code for trend extrapolations can be found in the ``Additional Analyses" vignette in the accompanying \href{https://github.com/agroimpacts/ecoscales}{R package/code repository}.

\subsubsection*{Extracting and analyzing data from earlier meta-analyses}
To compare the results of our analysis with the observational scales of earlier ecological studies, we used graph capture software\footnote{http://arohatgi.info/WebPlotDigitizer/app/} to extract the data values from Fig. 6.1 in Tilman \cite{tilman_ecological_1989}, Fig. 1 in Kareiva and Anderson \cite{kareiva_spatial_1988}, and Fig. 2 in Porter et al \cite{porter_crop_2005}. To maintain as much comparability as possible with our inclusion criteria, we excluded experiment studies in Tilman's data, as well as the values of any studies having duration values greater than 100 years (no upper time bound was provided for these), leaving duration values for 419 (out of 623) studies. Since Tilman presented duration values as a histogram, we calculated the mean duration across all studies as the weighted (by number of observations per bin) mean of bin center-point values (i.e. the weighted mean of the bin means).  We also excluded four (of 29) observation values from Porter et al's data on observational extent and frequency, which, in contrast to the other 25, were not randomly selected. Porter et al also used irregular scales for both their X (frequency) and Y (extent) axis, therefore we had to visually estimate the scale values for each data point after graphical extraction, and converted their extent values (in km) to hectares and their frequency values to intervals. Kareiva and Anderson presented resolution as plot diameters (m), which we squared to make comparable to our resolution metric.

Calculations of scale values from these studies can be found in the ``Additional Analyses" vignette in the accompanying \href{https://github.com/agroimpacts/ecoscales}{R package/code repository}.

@@

\subsubsection*{Trends in methods and scales}
\vspace{-10pt}
The results of weighted linear regression applied to observational types by year (Fig. \ref{type_by_yr}) suggest an increase (by 1.3\% per year) in the use of remote sensing and a corresponding decline in field methods (by the same percentage) during the 2004-2014 time period, although neither trend is statistically significant (remote observation regression, p$<$0.12; field observation regression, p$<$0.18).  Automated sensing methods showed no visible trend.

@@

Of course, we cannot attach confidence to a trend in observational extent extrapolated from a non-significant trend in observational type. However, some additional evidence in support of this possible trend lies within the extent values themselves. The regression analyses applied to dimensions shows an apparent increase in extent by 0.25 orders of magnitude per year (Fig. \ref{sc_by_yr}) between 2004-2014 (R$^2$ = 0.25, p$<$0.07). This somewhat clearer (albeit still not significant at the customary p$<$0.05 threshold) trend suggests that including more recent studies would have increased mean extent, but by a more modest 5.5\% (0.02 orders of magnitude).

### 2017-12-19 paper/main.tex
More edits to SI on trends section, and additional edits to main text

%This lower right to upper left orientation reflects the tradeoff between resolution and interval that is typical of satellite imaging \cite{estes_platform_2016}, and stands in contrast to the upper right to lower left line that stretches between this concentration and the high frequency (minute-hour intervals), high spatial resolution (0.1-100 cm$^2$) observations. This line demonstrates the opposite tradeoff that occurs with field-based observations, where larger plot sizes demand greater effort that in turn reduces sampling frequency \cite{kareiva_spatial_1988}.

@@

Comparing the interval and duration of temporally replicated observations showed most observations had daily to decadal intervals and durations of $\geq$1 month up to 1 decade (Fig. 2B). The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting one year to one decade tend to have yearly to decadal intervals. This tendency is reflected in the median domain locations of the primary observational methods: automated sensing had the finest median interval (hour-day) and shortest duration (month-year), followed by remote sensing (median interval slightly greater than one day and median duration of 1 year), field observations (median monthly interval and duration just over 1 year), and finally paleo-reconstructions (median interval 1 decade and millennial duration). The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken; this position

@@

\noindent \textbf{Discussion}\\
\noindent \textbf{\emph{Insights into the scale domains of modern ecology}}\\
Our results suggest that the scale domains of modern ecological observations are fairly narrow, and are collected primarily with conventional field techniques. Spatially, the majority of observations have grains of $\leq$1 m$^2$ and extents of $\leq$10,000 ha (Fig 1A;B). In the temporal domains, most observations are either un-replicated snapshots, or of low frequency ($\geq$1 month interval; Fig. 1C) and relatively short duration ($\leq$1 year; Fig. 1D; 2D). Contrasting observational dimensions reveals that larger extents are associated with larger plot sizes (Fig. 2C), while longer durations are associated with longer intervals (Fig. 2D). The latter association presumably reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation. The same tradeoff is also responsible for the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A). As a result of these tradeoffs, there are notable ``holes'' in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).

Have these domains changed since the seminal papers on scale first began to appear in the late 1980s \cite{wiens_spatial_1989, levin_problem_1992, tilman_ecological_1989}? A comprehensive answer to this question would require a similar study focused on earlier literature, but an analysis of results presented in three prior studies provides partial insight. The first and most comprehensive dataset consists of duration values extracted by Tilman \cite{tilman_ecological_1989} from 623 studies published between 1977-1987 in the journal \emph{Ecology}. The average duration of the most comparable subset of those values (n=419; see SI) was 3.6 years, compared to 3.3 years for observations in our sample (or 5.1 if temporally un-replicated observations are excluded). The second dataset is found in Kareiva and Anderson \cite{kareiva_spatial_1988}, who present the resolutions of 97 community ecology experiments published in \emph{Ecology} between 1980-1986. The average of those (12,657 m$^2$) was substantially smaller than the mean of our sample (1,496,069 m$^2$), but comparing the 80th percentile value (197 m$^2$) of Kareiva and Anderson's \cite{kareiva_spatial_1988} to ours (115 m$^2$) shows that the majority of contemporary observations are finer-grained than most 1980s-era experiments. The third dataset is provided by Porter et al \cite{porter_wireless_2005}, who compared the extent and interval of 25 studies published in 2003 and 2004 (also in \emph{Ecology}). The mean interval was 178 days, compared to 684 days in our sample, but the 80th percentile value in our study was 169 days compared to 329 days in theirs. Extent in our sample was substantially larger according to multiple summary statistics, including the mean (114,965,072 ha in our study versus 368,403 ha), median (5,051 ha versus 9 ha), and 95th percentile (46,424,808 ha versus 136,000 ha).

Although limited due to methodological differences (e.g. a focus on experiments versus un-manipulated systems), these comparisons suggest that the duration and, less clearly, resolution of ecological observations have changed little in the past 30 years, but observational frequency and particularly extent have increased. The growth in observational extent is also reflected by the positive year-on-year trend within our own dataset, which corresponds to the increasing use of remote sensing (see figures in SI).

However, even though observational extent is increasing, there remains for most observations a large gulf between the area that is actually sampled and that which the spatial replicates purportedly represent (Fig. 3A). A similarly large discrepancy is also evident between the time spanned by repeated observations and the time that is spent actually observing a phenomenon (Fig. 3B). These differences between the actual and \emph{representative} scales of observation have implications for ecological understanding, as the unobserved portions of space and time may contain important patterns and processes that are not captured by replicates, due to phenomenon-dependent factors such as autocorrelation and representativeness of the sampling scheme \cite{underwood_experiments_1997,palmer_scale_1994, cao_comparison_2002, legendre_spatial_1993,collins_method_2000}. Brief, infrequent snapshots, or fine-grained, spatially sparse replicates, may be sufficient to characterize many phenomena (as an example with respect to temporal replication, annual changes in tree cover are well-represented by low frequency satellite imaging \cite{hansen_high-resolution_2013}), but may be inadequate for more dynamic phenomena. For example, wildfire extent and duration can be mapped by daily return satellites \cite{roy_prototyping_2005,jones_fire_2009}, but the instantaneous nature of the imaging means that it cannot be used to observe fire behavior \cite{clements_observing_2007}. To capture such behavior, long periods of continuous observation may be more important for understanding dynamics than frequent repeats.

>>>>

@@

Comparing the interval and duration of temporally replicated observations showed most observations had daily to decadal intervals and durations of $\geq$1 month up to 1 decade (Fig. 2B). The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting 1 year to 1 decade tend to have yearly to decadal intervals. This tendency is reflected in the median domain locations of the primary observational methods: automated sensing had the finest median interval (hour-day) and shortest duration (month-year), followed by remote sensing (median interval slightly greater than one day and median duration of 1 year), field observations (median monthly interval and duration just over 1 year), and finally paleo-reconstructions (median interval 1 decade and millennial duration). The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken; this position

@@

\noindent \textbf{Discussion}\\
\noindent \textbf{\emph{Insights into the scale domains of modern ecology}}\\
Our results suggest that the scale domains of modern ecological observations are fairly narrow, and are collected primarily with conventional field techniques. Spatially, the majority of observations have grains of $\leq$1 m$^2$ and extents of $\leq$10,000 ha (Fig 1A;B). In the temporal domains, most observations are either un-replicated snapshots, or of low frequency ($\geq$1 month interval; Fig. 1C) and relatively short duration ($\leq$1 year; Fig. 1D).

Contrasting observational dimensions reveals that larger extents are associated with larger plot sizes (Fig. 2C), while longer durations tend to be associated with longer intervals (Fig. 2B). The latter association primarily reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation, the dominant observational mode, but is also reflected by the relative domain locations of the four main observational methods, suggesting that high observational frequency poses a durational cost across observational methods. A similar cost tradeoff is also evident in for the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A), which again primarily relates to characteristics of field-based observations, in which larger plot sizes demand greater effort that in turn reduces sampling frequency \cite{kareiva_spatial_1988}. Less obvious is the opposite tradeoff that occurs with remote observation (but shown in Fig. S2 in SI), where finer resolution is desired to increase detail, but typically comes at the cost of longer intervals \cite{estes_platform_2016}.

As a result of these aforementioned tradeoffs, there are several notable gaps in the spatio-temporal domains of ecological observations, which line in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).  Another hole is evident in the high to moderate resolution, large extent (1 million-10 billion ha) domain (Fig. 2C).

Have these domains changed since the seminal papers on scale first began to appear in the late 1980s \cite{wiens_spatial_1989, levin_problem_1992, tilman_ecological_1989}? A comprehensive answer to this question would require a similar study focused on earlier literature, but an analysis of results presented in three prior studies provides partial insight. The first and most comprehensive dataset consists of duration values extracted by Tilman \cite{tilman_ecological_1989} from 623 studies published between 1977-1987 in the journal \emph{Ecology}. The average duration of the most comparable subset of those values (n=419; see SI) was 3.6 years, compared to 3.3 years for observations in our sample (or 5.1 if temporally un-replicated observations are excluded). The second dataset is found in Kareiva and Anderson \cite{kareiva_spatial_1988}, who present the resolutions of 97 community ecology experiments published in \emph{Ecology} between 1980-1986. The average of those (12,657 m$^2$) was substantially smaller than the mean of our sample (1,496,070 m$^2$), but comparing the 80th percentile value (197 m$^2$) of Kareiva and Anderson's \cite{kareiva_spatial_1988} to ours (115 m$^2$) shows that the majority of contemporary observations are finer-grained than most 1980s-era experiments. The third dataset is provided by Porter et al \cite{porter_wireless_2005}, who compared the extent and interval of 25 studies published in 2003 and 2004 (also in \emph{Ecology}). The mean interval was 178 days, compared to 684 days in our sample, but the 80th percentile value in our study was 169 days compared to 329 days in theirs. Extent in our sample was substantially larger according to multiple summary statistics, including the mean (114,965,072 ha in our study versus 368,403 ha), median (5,051 ha versus 9 ha), and 95th percentile (46,424,808 ha versus 136,000 ha).

Although limited due to methodological differences (e.g. a focus on experiments versus un-manipulated systems), these comparisons suggest that the duration and, less clearly, resolution of ecological observations have changed little in the past 30 years, but observational frequency and particularly extent have increased. A weak positive trend in our own data suggests that average extent of ecological observations is steadily increasing (Fig. S5 in SI), and likely corresponds to the increasing use of remote observation technologies (Fig. S4 in SI).

However, even though observational extent appears to be increasing, there remains for most observations a large gulf between the area that is actually sampled and that which the spatial replicates purportedly represent (Fig. 3A). A similarly large discrepancy is also evident between the time spans covered by repeated observations and the time that is spent actually observing a phenomenon (Fig. 3B). These differences between the actual and \emph{representative} scales of observation have implications for ecological understanding, as the unobserved portions of space and time may contain important patterns and processes that are not captured by replicates, due to phenomenon-dependent factors such as autocorrelation and representativeness of the sampling scheme \cite{underwood_experiments_1997,palmer_scale_1994, cao_comparison_2002, legendre_spatial_1993,collins_method_2000}. Brief, infrequent snapshots, or fine-grained, spatially sparse replicates, may be sufficient to characterize many phenomena (as an example with respect to temporal replication, annual changes in tree cover are well-represented by low frequency satellite imaging \cite{hansen_high-resolution_2013}), but may be inadequate for more dynamic phenomena. For example, wildfire extent and duration can be mapped by daily return satellites \cite{roy_prototyping_2005,jones_fire_2009}, but the instantaneous nature of the imaging means that it cannot be used to observe fire behavior \cite{clements_observing_2007}. To capture such behavior, long periods of continuous observation may be more important for understanding dynamics than frequent repeats.
